---
title: "Semiparametric regression"
author: "Makowski Micha³"
date: "10 may 2017"
output:
  pdf_document:
    fig_caption: yes
    highlight: tango
    toc: yes
lang: en-EN
geometry: margin=1.5cm
subtitle: 'Homework Assignment #4'
fontsize: 10pt
---

```{r knitrOptions, include=FALSE}

inline_hook = function(x) { if (is.numeric(x)) { format(x, digits=2) } else x}

knitr::knit_hooks$set(inline=inline_hook)
knitr::opts_chunk$set(comment="", message=FALSE, tidy.opts=list(keep.blank.line=TRUE,
                      width.cutoff=120), options(width=100), warning=FALSE, error=FALSE, fig.align='center',
                      fig.height=5.5, fig.width=10, fig.show='hold',
                      size='footnotesize')

```

```{r libraries, include=FALSE}

rm(list=ls())

options(width=100)

# install.packages("HRW")
library(HRW)
# install.packages("mgcv")
library(mgcv)
# install.packages("lattice")
library(lattice)
# install.packages("nlme")
library(nlme)

data(femSBMD)
data(growthIndiana)

```

## Introduction

This couple pages cover fourth homework for Semiparametric Regression, 
a course conducted by proffesor Jaros³aw Harê¿lak at University of Wroc³aw. 
On the following pages three exercises will be presented. They are focused on mixed models 
on grouped data. Examples of such data are obtain during medical studies in which patients 
are followed over time and measurements on them recorded repeatedly, 
educational studies in which students grouped into classrooms and schools are scored 
on examinations and sample surveys in which the respondents to questionnaires are 
grouped within geographical districts.

Mixed models are a good choice for the analysis of grouped data, with random effects used
to account for within-group dependence. 

Packages __HRW__ (data), __mgcv__ (model fitting) and __lattice__ (visualization)
will be used.

\newpage
## Exercise II (Additive Mixed Models)

In exercise II we want to fit smooth line to the data about spinal bone mineral density.

At first, we plot the data divided by ethnic group.

```{r 2_intro}

xyplot(spnbmd~age|factor(ethnicity), 
       group = idnum, 
       data = femSBMD, 
       xlab = "age (years)", 
       ylab = "spinal bone mineral density (g/cm2)", 
       panel = function(x,y,subscripts,groups)
       {
           panel.grid()
           panel.superpose(x,y,subscripts,groups,type = "b",pch = 16,lwd = 2)
       })

```

Above graph shows how the spinal bone mineral density chages during the life of a single
represantative in each cohort. We can see some dependecies, which are different in each group 
and we would like to focus on them a little more.

For such data we will use following model:
$$
\text{spnbmd}_{i,j}=U_i+f(\text{age}_{i,j})+\beta_1 \text{black}_i+\beta_2 \text{hispanic}_i+\beta_3 \text{white}_i+\epsilon_{i,j}\\
1\le j\le n_i,\quad 1\le i\le 230\\
U\sim N(0,\sigma^2_U),\epsilon_i\sim N(0,\sigma^2_{\epsilon})
$$
where $\text{black}_i$, $\text{hispanic}_i$, $\text{white}_i$ are of course indicators
and $\text{spnbmd}_{i,j}$ are the mesurments of spinal bone density 
of *i*th subject at *j*th time.

\newpage
We could interprate this model as following: Asian group is the "basis", the reference, the coefficients
$\beta_1$, $\beta_2$ and $\beta_3$ are the modifications of the mean in other ethics groups.

```{r 2_model}

fit = gamm(spnbmd~s(age) + black + hispanic + white, random = list(idnum = ~1), data = femSBMD)
plot(fit$gam, shade = TRUE, shade.col = "palegreen")

```

The shaded region corresponds to pointwise approximate 95% confidence in-
tervals. Note that default plotting of the estimate of $f(\text{age})$ involves vertical centering about zero.

\newpage
Let's have a look at the summary

```{r 2_summary}

summary(fit$gam)

```

Note that the fitted age effect involves 7.201 effective degrees of freedom.
Let's look at the confidence intervals

```{r 2_intervals}

intervals(fit$lme)

```

This output shows that an approximate 95% confidence interval for $\beta_1$ is $(0.0482,0.116)$,
which indicates a statistically significant difference between the Asian and Black
females in terms of mean spinal bone mineral density. However, there is no
significant difference between Hispanic or White females and Asian females.
An approximate 95% confidence interval for $\sigma_{U}$ is $(0.114,0.131)$, which 
implies significant within-subject correlation. The 95% confidence interval for
$\sigma_{\epsilon}$ is $(0.0347,0.0390)$.

\newpage
## Exercise III (Models with Group-Specific Curves)

In exercise III we want to fit group-specific model to the data of adolescent 
somatic growth obtained from a study of the mechanisms of human hypertension 
development conducted at the Indiana University School of Medicine, Indianapolis, 
Indiana, USA. Pratt et al.(1989).

We restrict attention to the black males in the study. The following plot show the trend of 
each subject, there are 28 of them.

```{r 3_intro}

growthINblackMales = growthIndiana[(growthIndiana$male == 1) & (growthIndiana$black == 1),]

xyplot(height~age|idnum,
       group = idnum,
       data = growthINblackMales,
       layout = c(4,7),
       strip = F,
       xlab = "age (years)",
       ylab = "height (centimeters)",
       as.table = TRUE,
       panel = function(x,y,subscripts,groups)
       {
           panel.grid()
           panel.superpose(x,y,subscripts,groups,col = "dodgerblue",type = "b")
       })

```

Above graph shows the history for each, individual representative. We notice that 
he shapes of the curves for each adolescent differ quite markedly
and the simple additive mixed models would not capture such behavior very well.

Instead, such data we will use following model:
$$
\text{height}_{i,j}=f(\text{age}_{i,j})+g_i(\text{age}_{i,j})+\epsilon_{i,j}\\
1\le j\le n_i,\quad 1\le i\le 28\\
\epsilon_{i}\sim N(0,\sigma^2_{\epsilon})
$$
where $n_i$ is the number of measurments for *i*th subject, $g_i$ is a function
that represents that adolescent’s departure from the overall mean function $f$.
This is **semiparametric mixed model**.

\newpage
Modeling of $f$ and the $g_i$ can proceed according to

$$
f(x) = \beta_0 + \beta_1 x + \sum_{k=1}^{K_{gbl}}u_{gbl,k}z_{gbl,k}(x), 
\qquad u_{gbl,k}|\sigma_{gbl}\sim N(0,\sigma^2_{gbl})\\
g_i(x)=U_{0i}+U_{1i}x+\sum_{k=1}^{K_{grp}}u_{grp,ik}z_{grp,k}(x)\\
\begin{bmatrix}
    U_{0i}\\
    U_{1i}
\end{bmatrix}
\sim N(0,\Sigma), \qquad u_{grp,ik}|\sigma_{grp}\sim N(0,\sigma^2_{grp})
$$

where $z_{gbl,k}$ and $z_{grp,k}$ are suitable spline bases of sizes
$K_{gbl}$ and $K_{grp}$ respectively. We will use canonical O’Sullivan splines. 
Typically, $K_{grp}$ is smaller than $K_{gbl}$ since
fewer basis functions are needed to handle group-specific deviations.

\newpage
## Exercise IV (Fitting Group-Specific Curves Models)

To fit model introduced in exercise III we will use package **nlme**.

At first we wxtract the necessary variables from the __growthINblackMales__
dataset

```{r 4_dataPrep}

age = growthINblackMales$age
height = growthINblackMales$height
idnum = growthINblackMales$idnum

```

Then we create an array with identification numbers

```{r 4_dataPrep2}

uqID = unique(idnum)
uqID.tab = table(idnum)
uqID.len = length(uqID)
growthINblackMales$idnumBM = as.numeric(factor(rep(uqID, uqID.tab), labels= 1:uqID.len))
idnumBM = growthINblackMales$idnumBM

```

Next, we will set up the design matrices $Z_{gbl}$, containing the
$Z_{gbl,k}$, and $Z_{grp}$, containing the $Z_{grp,k}$

```{r 4_matrices1}

numObs = length(height)
numGrp = uqID.len
numIntKnotsGbl = 20
intKnotsGbl = quantile(unique(age),
seq(0,1,length=numIntKnotsGbl+2))[-c(1,numIntKnotsGbl+2)]
range.age = c(5.5,20)
Zgbl = ZOSull(age,range.x=range.age,intKnots=intKnotsGbl)

numIntKnotsGrp = 10
intKnotsGrp = quantile(unique(age),
seq(0,1,length=numIntKnotsGrp+2))[-c(1,numIntKnotsGrp+2)]
Zgrp = ZOSull(age,range.x=range.age,intKnots=intKnotsGrp)

```

Then we set up the random effect structure for the call to $\text{lme()}$

```{r 4_matrices2}

dummyId = factor(rep(1,numObs))
Zblock = list(dummyId=pdIdent(~-1+Zgbl),idnumBM=pdSymm(~age),idnumBM=pdIdent(~-1+Zgrp))

```

Let have a brief look at operation done above:


* The dummy identification variable $\text{dummyID}$, an array of length $\text{numObs}$, the total number of observations, with all entries equal to one tricks $\text{lme()}$ into accommodating the global penalized spline component.
* The list entry $\text{dummyId=pdIdent(~-1+Zgbl)}$ invokes the multiple of identity matrix structure $u_{gbl}\sim N(0,\sigma^2_{gbl}I)$ across the entire dataset regardless of within-subject grouping.
* The list item $\text{idnumBM=pdSymm(~age)}$ invokes the block-diagonal unstructured $2\times2$ covariance matrix form on the $[U_{0i},U_{1i}]^T$, $1\le i\le 23$, as required by the model, 
* Similarly $\text{pdIdent(~-1+Zgrp)}$ accommodates $u_{grp,ik}|\sigma_{grp}\sim N(0,\sigma^2_{grp})$

We are now ready to call $\text{lme()}$ with the random argument set to $\text{Zblock}$

```{r 4_mode}

blkMalGD = groupedData(height ~ age|rep(1,length = numObs), 
                        data = data.frame(height,age,Zgbl,Zgrp,idnumBM))
fit = lme(height ~ age,data = blkMalGD,random = Zblock)

```

Now let's plot fitted curves, firtsly we wetup the fixed and random design matrices

```{r 4_plotting}

ng = 201
ageg = seq(range.age[1],range.age[2],length = ng)
Xg = cbind(rep(1,ng),ageg)
Zgblg = ZOSull(ageg,range.x = range.age,
intKnots = intKnotsGbl)
Zgrpg = ZOSull(ageg,range.x = range.age,
intKnots = intKnotsGrp)

```

\newpage
Then we extract the model coefficients

```{r 4_extractingCoef}

betaHat = as.vector(fit$coef$fixed)
uHat = as.vector(fit$coef$random[[1]])
fHatg = as.vector(Xg%*%betaHat + Zgblg%*%uHat)

```

In the end we estimate the subject-specific curves

```{r 4_estCurves}

curvEsts = vector("list",numGrp)
for (i in 1:numGrp)
{
    uLinHati = as.vector(fit$coef$random[[2]][i,])
    uSplHati = as.vector(fit$coef$random[[3]][i,])
    ghati = Xg%*%uLinHati + Zgrpg%*%uSplHati
    curvEsts[[i]] = fHatg + ghati
}

```

Finally, we are ready to plot the curves

```{r 4_plot}

xyplot(height ~ age|idnumBM,groups = idnumBM,
       data = growthINblackMales,
       strip = FALSE,
       xlab = "age (years)",
       ylab = "height (centimeters)",
       as.table = TRUE, layout = c(4,7),
       panel = function(x,y,subscripts,groups)
       {  
           panel.grid()
           adolNum = idnumBM[subscripts][1]
           panel.superpose(x,y,subscripts,groups,col = "dodgerblue",type = "b")
           panel.xyplot(ageg,curvEsts[[adolNum]],col = "blue",type = "l")
       })

```

\newpage
We have to remember to check the standardized residuals from the fit shown in the
last plot. Let's see

```{r 4_plotFit}

plot(fit)

```

We connot see any distinct patterns or outliers.

\newpage
## Exercise V 

Model presented in **Exercise II** for the female spinal bone mineral density data assumes that the mean
functions for each ethnicity category differ only by vertical shifts.
A more flexible model is

$$
\text{spnbmd}_{i,j}=U_i+f_{ethnicity}(\text{age}_{i,j})+\epsilon_{i,j}\\
1\le j\le n_i,\quad 1\le i\le 230\\
U\sim N(0,\sigma^2_U),\epsilon_i\sim N(0,\sigma^2_{\epsilon})
$$
where
$$
\text{etnicity}\in\{\text{asian},\text{black},\text{hispanic},\text{white}\},\quad 1\le i\le 230.
$$

Now we will create such model and look at it's summary

```{r 5_model}

fitEthno = gamm(spnbmd~s(age, by=ethnicity), random = list(idnum = ~1), data = femSBMD)
summary(fitEthno$gam)

```

We can see that smoothing splines for Asian and Hispanic are quite similiar, while for other ethnics groups 
they differ. That correlate with our summory of **Exercise II**. Now we will plot the smooting splines for each group. For more analysis we could call $\text{summary(fitEthno\$gam)}$, it gives us more information 
about mixed effects.


\newpage
Let's compare fitted splines graphicaly 
```{r 5_sumamry}

plot(fitEthno$gam, shade = TRUE, shade.col = "palegreen", pages = 2)

```

All plots slightly differ from each other. The axes are locked for every group, 
that let us to compare each other.







