---
title: "Semiparametric regression"
author: "Makowski Micha³"
date: "10 may 2017"
output: 
  pdf_document: 
    fig_caption: true
    highlight: tango
    toc: true
fontsize: 10pt
geometry: margin=1.5cm
subtitle: 'Homework Assignment #3' 
lang: en-EN 
---

```{r knitrOptions, include=FALSE}

options(scipen = 999)

inline_hook = function(x) { if (is.numeric(x)) { format(x, digits=2) } else x}

knitr::knit_hooks$set(inline=inline_hook)
knitr::opts_chunk$set(fit.align="center", warning=FALSE, error=FALSE, message=FALSE, echo=FALSE,
                      tidy.opts=list(keep.blank.line=TRUE, width.cutoff=120),
                      options(width=100), fig.height=5.5,
                      fig.width=10, fig.show='hold', size='footnotesize')

```

```{r libraries, include=FALSE}

rm(list=ls())

options(width=100)
library(gss)
library(gam)
library(lattice)
library(HRW)
library(GGally)

```

# Introduction

This couple pages cover third homework for Semiparametric Regression, 
a course conducted by proffesor Jaros³aw Harê¿lak at University of Wroc³aw. 
On the following pages two exercises will be presented. First one compares two different methods of fitting splines on data about 
ozone concentration in Los Angeles. Second is focused on comparing diffrent curve fits on data about milk characteristc among
different cows with different diets.

\newpage

# Exercise I

In this exercise we want to construct a model, which says us how ozone concentration depends on others variables.
To do it we use Generalized Additive Models.

## Description of data

Daily measurements of ozone concentration and eight meteorological quantities in the Los Angeles basin for 330 days of 1976.

A data frame containing 330 observations on the following variables.

* **upo3**-	 Upland ozone concentration, in ppm.  
* **vdht**-	 Vandenberg 500 millibar height, in meters.  
* **wdsp**-	 Wind speed, in miles per hour.  
* **hmdt**-	 Humidity.  
* **sbtp**-	 Sandburg Air Base temperature, in Celsius.  
* **ibht**-	 Inversion base height, in foot.  
* **dgpg**-	 Dagget pressure gradient, in mmHg.  
* **ibtp**-	 Inversion base temperature, in Fahrenheit.  
* **vsty**-	 Visibility, in miles.  
* **day**-   Calendar day, between 1 and 366. 

## Relations beetween data

Let's invistigate data with `ggpairs(ozone)` command.

```{r ggPairs1}

data(ozone)
ggpairs(ozone)

```

We could see that relations are mostly non linear. 

## Model selection

We build full model and then, using *step.Gam()* function, we to select predictors and their diffrent forms among Gaussian GAMs 
with `upo3` as the response variable. We will choose beetween linear and nonlinear forms.

```{r gamStep1}

fitUpo3 <- gam(upo3 ~ vdht + wdsp + hmdt + sbtp + ibht + dgpg + ibtp + vsty + day, 
               data = ozone)

stepFitupo3 <- step.Gam(fitUpo3,
                        scope = list("vdht" = ~ 1+vdht+s(vdht,2),
                                     "wdsp" = ~ 1+wdsp+s(wdsp,2),
                                     "hmdt" = ~ 1+hmdt+s(hmdt,2),
                                     "sbtp" = ~ 1+sbtp+s(sbtp,2),
                                     "ibht" = ~ 1+ibht+s(ibht,2),
                                     "dgpg" = ~ 1+dgpg+s(dgpg,2),
                                     "ibtp" = ~ 1+ibtp+s(ibtp,2),
                                     "vsty" = ~ 1+vsty+s(vsty,2),
                                     "day"  = ~ 1+day +s(day,2)))

```

For stepwise search AIC is the smallest for model with predictors presented below.

```{r modelSummary1, echo=TRUE}

names(stepFitupo3$"model")[-1]

```

## Model with GCV-based smoothing parameter selection

Now we fit model with predictors, which was chosen in previous part.

```{r libraryDetach1}

detach(package:gam)
library(mgcv)
apply(ozone, 2, function(x){length(unique(x))})

```

Model summary is presented below.

```{r mgcvModel1, echo=TRUE}

fitStepOzone <- gam(upo3 ~ vdht + wdsp + hmdt +
                        s(sbtp, k=2) + s(ibht, k=2) + s(dgpg, k=2) + 
                        s(ibtp, k=2) + s(vsty, k=2) + s(day, k=2), data = ozone)

fitStepOzone

```

## Model modyfication

In this part we want to choose model with sufficient number of basis functions. 
To do this we loop over a number of basis function level until we receive satisfactory results.
We have to modify k's manually - chosen formula is presented below.
There should be more diagnostics done in this part to develop best model and fully check it.

```{r loop1}

m <- 20
p <- vector(length = m-2)

for( i in 3:m)
{
    fitStepOzoneTemp <- gam(upo3 ~ vdht + wdsp + hmdt +
                                s(sbtp, k=i) + 
                                s(ibht, k=i) +
                                s(dgpg, k=i) + 
                                s(ibtp, k=i) +
                                s(vsty, k=i) + 
                                s(day, k=i), data = ozone)
}

# plot(3:m, p, xlab = "k-index", ylab = "p-value") 
# lines(3:m, rep(0.05, length.out=m-2), col = "red")

fitStepOzoneTemp6 <- gam(upo3 ~  
                             vdht  + wdsp + hmdt + 
                             s(sbtp, k=5) + s(ibht, k=7) + s(dgpg, k=15) + 
                             s(ibtp, k=25) + s(vsty, k=5) + s(day, k=15),  
                         data = ozone)


# summary(fitStepOzoneTemp6)$s.pv

```



```{r residualsCheck1}

gam.check(fitStepOzoneTemp6)

```

Probability distribution of deviance residuals is almost normal, but the residuals variance is not constant. 
There is a very quadratic dependence between residuals and linear predictors, 
Unfortunetelly, we were unable to find model, which looks MUCH better than this presented above.

\newpage
# Exercise II

In this task we have data contaning measurements of the milk protein level over time since calving. 
Data contains information for of 79 cows. There were three types of diet.
In first part we illustrate the fitting of additive mixed models for all cows. 
Next we show group-specific curves semiparametric mixed model for each diet group.

First we, we will dive into the data more.

```{r milkIntro2}

library(dplyr)

data("Milk")


Milk %>% 
        mutate( barley = as.numeric( (Diet == "barley") ),
                lupins = as.numeric( (Diet == "lupins"))) ->
        Milk

summary(Milk)

```

```{r plot2}

library(ggplot2)

ggplot(Milk, aes(x = Time, y = protein, color = Cow)) +
    geom_point() +
    geom_line() +
    guides(color=FALSE) + 
    facet_grid(Diet~ .) +
    labs(title = "Protein content of milk",
         subtitle= "Grouped by diet",
         x = "Time moment") +
    scale_x_continuous(labels = seq(0,20,2),
                       breaks = seq(0,20,2)) +
    theme_minimal()

```

## GAMM

Now we build the first model. 
We have 79 cows and three groups: **barley, lupins** and **barley+lupins**. 

$$ protein_{ij} = U_i + f(time_{ij}) +\beta_1barley_i+\beta_2 lupins+\epsilon_{ij}$$
$$U_i \overset{iid}{\sim} \mathcal{N}(0,\sigma^2), \qquad \epsilon_{ij}\overset{iid}{\sim} \mathcal{N}(0,\sigma_{\epsilon}^2) $$
We fit GAMM with fixed **time**, **diet group** and random subject **cow number*.*

```{r firstFit2}

fitProt <- gamm(protein ~ s(Time)+lupins+barley, random = list(Cow = ~1), data = Milk )

```

Now we plot and sum up centralized penelized spline.

```{r plotModel2}

plot(fitProt$gam, shade = T, shade.col = "palegreen")

summary(fitProt$gam)

```

We could observe low $R^2$. 
All predictors are useful regarding t-test

## Group specific splines

Here we focus on group-specific curvs 

As we could see below models seems to be rather well fitted, but there are some outlier visible.

```{r dplyr2}

ExtractDiet <- function(name_diet)
{
        Milk %>%
        filter( Diet == name_diet )  
}

Barley <- ExtractDiet("barley") 
Lupins <- ExtractDiet("lupins") 
BarLup <- ExtractDiet("barley+lupins") 

```

```{r, echo = F, message = T}

MilkData <- Barley
dimplot <- c(5,5)
title <- "Barley"

        MilkData$Cow <- as.vector(MilkData$Cow) 
        time <- MilkData$Time
        protein <- MilkData$protein
        cow <- MilkData$Cow
        
        uqCow <- unique(cow)
        uqCowTab <- table(cow)
        uqCowLen <- length(uqCow)
        MilkData$cowBM <- as.numeric( factor(rep(uqCow,uqCowTab),
                                             labels = 1:uqCowLen))
        cowBM <- MilkData$cowBM
        
        numGrp <- uqCowLen
        numIntKnotsGbl <- 20
        intKnotsGbl <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGbl+2))[-c(1,numIntKnotsGbl+2)] 
        
        range.time <- c(min(MilkData$Time), max(Milk$Time))
        Zgbl <- ZOSull(time, range.x = range.time, intKnots = intKnotsGbl)
        
        numIntKnotsGrp <- 10
        intKnotsGrp <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGrp+2))[-c(1,numIntKnotsGrp+2)]
        Zgrp <- ZOSull(time, range.x = range.time, intKnots = intKnotsGrp)
        
        dummyId <- factor(rep(1,length(protein)))
        
        Zblock <- list(dummyId = pdIdent(~ -1 + Zgbl),
                       cowBM = pdSymm(~time),
                       cowBM = pdIdent(~-1+Zgrp))
        
        
        MilkGD <- groupedData(protein ~ time|rep(1, length = length(protein)),
                              data = data.frame(protein, time, Zgbl, Zgrp, cowBM ))
        fitMilkGD <- lme(protein ~ time, data=MilkGD, random = Zblock)

        ng <- 250
        timeg <- seq(range.time[1], range.time[2], length = ng)
        Xg <- cbind(rep(1,ng), timeg)
        Zgblg <- ZOSull(timeg,range.x = range.time, intKnots = intKnotsGbl)
        Zgrpg <- ZOSull(timeg,range.x = range.time, intKnots = intKnotsGrp)
        
        betaHat <- as.vector(fitMilkGD$coef$fixed)
        uHat <- as.vector(fitMilkGD$coef$random[[1]])
        fHatg <- as.vector(Xg%*%betaHat + Zgblg%*%uHat)
        
        curvEsts <- vector("list", numGrp)
        for(i in 1:numGrp){
                uLinHati <- as.vector(fitMilkGD$coef$random[[2]][i,])
                uSplHati <- as.vector(fitMilkGD$coef$random[[3]][i,])
                ghati <- Xg%*%uLinHati + Zgrpg%*%uSplHati
                curvEsts[[i]] <- fHatg + ghati
        }
        
        xyplot( protein ~ time|cowBM, groups = cowBM,
                data = MilkData,
                strip = FALSE,
                xlab = "Time moment",
                ylab = "Protein level",
                main = title,
                as.table = TRUE, layout = c(dimplot[1],dimplot[2]),
                panel = function(x,y,subscripts,groups){
                        
                        panel.grid()
                        adolNum <- cowBM[subscripts][1]
                        panel.superpose(x,y,subscripts,groups,
                        col = "darkgreen",type = "b")
                        panel.xyplot(timeg, curvEsts[[adolNum]],
                                     col = "palegreen", type ="l")
                })

```

```{r, echo = F, message = T}

MilkData <- Lupins
dimplot <- c(6,5)
title <- "Lupin"

        MilkData$Cow <- as.vector(MilkData$Cow) 
        time <- MilkData$Time
        protein <- MilkData$protein
        cow <- MilkData$Cow
        
        uqCow <- unique(cow)
        uqCowTab <- table(cow)
        uqCowLen <- length(uqCow)
        MilkData$cowBM <- as.numeric( factor(rep(uqCow,uqCowTab),
                                             labels = 1:uqCowLen))
        cowBM <- MilkData$cowBM
        
        numGrp <- uqCowLen
        numIntKnotsGbl <- 20
        intKnotsGbl <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGbl+2))[-c(1,numIntKnotsGbl+2)] 
        
        range.time <- c(min(MilkData$Time), max(Milk$Time))
        Zgbl <- ZOSull(time, range.x = range.time, intKnots = intKnotsGbl)
        
        numIntKnotsGbl <- 20
        intKnotsGbl <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGbl+2))[-c(1,numIntKnotsGbl+2)] 
        
        range.time <- c(min(MilkData$Time), max(Milk$Time))
        Zgbl <- ZOSull(time, range.x = range.time, intKnots = intKnotsGbl)
        
        numIntKnotsGrp <- 10
        intKnotsGrp <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGrp+2))[-c(1,numIntKnotsGrp+2)]
        Zgrp <- ZOSull(time, range.x = range.time, intKnots = intKnotsGrp)
        
        
        dummyId <- factor(rep(1,length(protein)))
        
        Zblock <- list(dummyId = pdIdent(~ -1 + Zgbl),
                       cowBM = pdSymm(~time),
                       cowBM = pdIdent(~-1+Zgrp))
        
        MilkGD <- groupedData( protein ~ time|rep(1, length = length(protein)),
                               data = data.frame(protein, time, Zgbl, Zgrp, cowBM ))
        fitMilkGD <- lme(protein ~ time, data=MilkGD, random = Zblock)

        ng <- 250
        timeg <- seq(range.time[1], range.time[2], length = ng)
        Xg <- cbind(rep(1,ng), timeg)
        Zgblg <- ZOSull(timeg,range.x = range.time, intKnots = intKnotsGbl)
        Zgrpg <- ZOSull(timeg,range.x = range.time, intKnots = intKnotsGrp)
        
        betaHat <- as.vector(fitMilkGD$coef$fixed)
        uHat <- as.vector(fitMilkGD$coef$random[[1]])
        fHatg <- as.vector(Xg%*%betaHat + Zgblg%*%uHat)
        
        curvEsts <- vector("list", numGrp)
        for(i in 1:numGrp){
                uLinHati <- as.vector(fitMilkGD$coef$random[[2]][i,])
                uSplHati <- as.vector(fitMilkGD$coef$random[[3]][i,])
                ghati <- Xg%*%uLinHati + Zgrpg%*%uSplHati
                curvEsts[[i]] <- fHatg + ghati
        }
        
        xyplot( protein ~ time|cowBM, groups = cowBM,
                data = MilkData,
                strip = FALSE,
                xlab = "Time moment",
                ylab = "Protein level",
                main = title,
                as.table = TRUE, layout = c(dimplot[1],dimplot[2]),
                panel = function(x,y,subscripts,groups){
                        
                        panel.grid()
                        adolNum <- cowBM[subscripts][1]
                        panel.superpose(x,y,subscripts,groups,
                        col = "darkgreen",type = "b")
                        panel.xyplot(timeg, curvEsts[[adolNum]],
                                     col = "palegreen", type ="l")
                })

```



```{r, echo = F, message = T}
MilkData <- BarLup
dimplot <- c(6,5)
title <- "Barley + lupin"

        
        MilkData$Cow <- as.vector(MilkData$Cow) 
        time <- MilkData$Time
        protein <- MilkData$protein
        cow <- MilkData$Cow
        
        uqCow <- unique(cow)
        uqCowTab <- table(cow)
        uqCowLen <- length(uqCow)
        MilkData$cowBM <- as.numeric( factor(rep(uqCow,uqCowTab),
                                             labels = 1:uqCowLen))
        cowBM <- MilkData$cowBM
        
        numGrp <- uqCowLen
        numIntKnotsGbl <- 20
        intKnotsGbl <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGbl+2))[-c(1,numIntKnotsGbl+2)] 
        
        range.time <- c(min(MilkData$Time), max(Milk$Time))
        Zgbl <- ZOSull(time, range.x = range.time, intKnots = intKnotsGbl)
        
        numIntKnotsGbl <- 20
        intKnotsGbl <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGbl+2))[-c(1,numIntKnotsGbl+2)] 
        
        range.time <- c(min(MilkData$Time), max(Milk$Time))
        Zgbl <- ZOSull(time, range.x = range.time, intKnots = intKnotsGbl)
        
        numIntKnotsGrp <- 10
        intKnotsGrp <- quantile(unique(time),
                        seq(0,1,length=numIntKnotsGrp+2))[-c(1,numIntKnotsGrp+2)]
        Zgrp <- ZOSull(time, range.x = range.time, intKnots = intKnotsGrp)
        
        
        dummyId <- factor(rep(1,length(protein)))
        
        Zblock <- list(dummyId = pdIdent(~ -1 + Zgbl),
                       cowBM = pdSymm(~time),
                       cowBM = pdIdent(~-1+Zgrp))
        
        
        MilkGD <- groupedData( protein ~ time|rep(1, length = length(protein)),
                               data = data.frame(protein, time, Zgbl, Zgrp, cowBM ))
        fitMilkGD <- lme(protein ~ time, data=MilkGD, random = Zblock)

        ng <- 250
        timeg <- seq(range.time[1], range.time[2], length = ng)
        Xg <- cbind(rep(1,ng), timeg)
        Zgblg <- ZOSull(timeg,range.x = range.time, intKnots = intKnotsGbl)
        Zgrpg <- ZOSull(timeg,range.x = range.time, intKnots = intKnotsGrp)
        
        betaHat <- as.vector(fitMilkGD$coef$fixed)
        uHat <- as.vector(fitMilkGD$coef$random[[1]])
        fHatg <- as.vector(Xg%*%betaHat + Zgblg%*%uHat)
        
        curvEsts <- vector("list", numGrp)
        for(i in 1:numGrp){
                uLinHati <- as.vector(fitMilkGD$coef$random[[2]][i,])
                uSplHati <- as.vector(fitMilkGD$coef$random[[3]][i,])
                ghati <- Xg%*%uLinHati + Zgrpg%*%uSplHati
                curvEsts[[i]] <- fHatg + ghati
        }
        

        xyplot( protein ~ time|cowBM, groups = cowBM,
                data = MilkData,
                strip = FALSE,
                xlab = "Time moment",
                ylab = "Protein level",
                main = title,
                as.table = TRUE, layout = c(dimplot[1],dimplot[2]),
                panel = function(x,y,subscripts,groups)
                {  
                   panel.grid()
                        adolNum <- cowBM[subscripts][1]
                   panel.superpose(x,y,subscripts,groups,                        
                                   col = "darkgreen",type = "b")
                   panel.xyplot(timeg, curvEsts[[adolNum]],
                                col = "palegreen", type ="l")
                })
        

```